\thesistitle

### 2025-07-24 manuscript/part2.tex
Acknowl. Contentspage, pagenumbering for pretexts
slight fixes; stremline part2

As discussed in in the introduction \ref{sec:scope}, the main aim is the development of a VS Code Language extension that operates on par with
well-established tooling for other programming languages. The following requirements define the expected capabilities and characteristics the extension must demonstrate.

@@

  \subsection*{Non-Functional Requirements}
  \begin{enumerate}
    \item The extension must respond to user edits with syntax and semantic feedback within 100-250ms for files of average size.
    \item The extension must provide a user-friendly experience that is consistent with other popular VS Code extensions.
    \item The extension must be structured and commented in a way that allows future developers to understand and extend its functionality easily.
    \item The extension must be compatible with the latest version of Visual Studio Code at the time of its release and be developed using Langium, with consideration for future maintenance and version support.
    \item The extension must be able to handle files of immense size (around 10000 lines) of code within the bounds of 1 second.
    \item The extension must be delivered with clear and comprehensive documentation, including installation, usage, and extension of its features.
  \end{enumerate}
\end{multicols}

\section{Alternative Technologies}

This extension is implemented using \textbf{Langium}, which itself builds upon the standardized \textbf{Language Server Protocol (LSP)} \cite{LSP}, yet
there are several other technologies and approaches available for creating Visual Studio Code language extensions. At its core, the LSP allows tooling for programming
languages to be developed in a modular, editor-independent manner, making it possible to implement a language server in any language that can communicate via JSON-RPC.

@@

For basic syntax highlighting and lexical analysis, \textbf{TextMate}-style grammars~\cite{TextMate} can be used, offering a lightweight approach when deep semantic understanding is not required.
Meanwhile, incremental parsing frameworks such as \textbf{Tree-sitter}~\cite{Treesitter} have gained popularity for their ability to provide high-performance syntax parsing and error recovery,
making them ideal for tooling that operates on very large files.

@@

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as
    the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
    from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
    All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
    which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textsf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
    dependency injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
    tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
    state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
    Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possible. Moreover, Langium enables developers to swap in
    custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec:customservices}.
    By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
    However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

Both \code{module} files share analogous responsibilities by encapsulating dependency-injection setup and service registration for the language environment.
The generated module is produced automatically from the grammar definition and offers two DI fragments, one that registers shared-core services such as AST
reflection and another that configures the parser, the Chevrotain grammar factory and language metadata. In contrast the main module is handwritten and imports
Langium's default core and shared fragments, registers any custom services such as validators, scope providers and caches, invokes \code{inject} to assemble the full
service container and exposes a \code{createServices} factory.

@@

\subsection*{Language Server Protocol: Server client interaction}

This illustration is a concise comprehension aid to show how the Language Server Protocol enables standardized communication between an editor (client) and a language server:
from the initialization handshake through document events, feature requests (completion, hover, definition, references, formatting, semantic tokens), and shutdown \cite{lspFeatures}.
By defining a common set of well-specified messages and payloads, LSP serves as a generally applicable protocol that any editor and any language server can adopt to instantly
share rich tooling capabilities without per-language or per-IDE custom integrations.

@@

A Langium grammar definition primarily consists of rules: \code{<Rulename>: <production_result>} and terminals: \code{terminal <name>: <result>} defining the language
in accordance of conventional grammatical specifications. In addition, it must also include keywords specifying the grammar name also used in the AST: \code{grammar <name>},
and to determine the start from which a document should be parsed: \code{entry <RootRulename>: <production_result>}.
In this case, the production result describes the whole parsable document and therefore acts as the root node of the AST.
\\
Also mention the LL (2) parsability and which tokens demand a 2 lookahead ()
The code excerpts in the following sections are all taken from the original MiniProb Langium grammar (\ref{list:langium-grammar}) and, for easier digestion,
have been truncated; each snippet focuses respectively on the core rules, the expression system, and the terminal and type definitions.
\subsection*{Core rules}

The code excerpt shown here defines 'MiniProb' as the grammar name and sets the \code{Program} rule as the entry point, adhering to nominal abstraction by
referring to each MiniProb-parsed document as a program. Lines of interest include line 8 - showing that every MiniProb program consists of some global declarations
followed by at least on function - and line 4, which allows for file import of other MiniProb programs at the beginning of the document.

The \code{FileImport} rule emulates the C preprocessor's "\#include" directive by requiring a literal match of the \code{#include} keyword followed by a
file name token (terminal \code{STRING}). Generally during parsing, each line is evaluated either against these fixed string keywords or against any matching terminals (order of match).
As the file names have to be retrievable once it is time to compute the scope \ref{sec:customservices}, they are made accessible through the AST
, with each FileImport-node containing a \code{file} property that stores the corresponding value.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos, breaklines]{yaml}
grammar MiniProb
type DeclOrParam = Decl | Param;
entry Program:
    (fileImports+=FileImport)*
    probabilisticQuery=PROB_QUERY? // not tested: out of scope
    formula=FORMULA? // not tested: out of scope
    'program:'?
    (declarations+=Decl ';')* functions+=Func+;
FileImport:
    '#include' file=STRING;
Decl:
    type=Type names+=ID (',' names+=ID)*;
.
.
Lval:
    ref=[DeclOrParam:ID] ('['index=Expression']')?;
.
.
Param:
    type=Type byRef?='&'? name=ID;
ArgList:
    arguments+=Arg (',' arguments+=Arg)*;
Arg:
    expression=Expression; //only expression and no by-ref symbol (could be replaced)
.
.
\end{minted}
\captionof{listing}{Langium grammar core of MiniProb}

@@

Most of the expression system is omitted here, as it closely follows the \textit{operator-precedence grammar} pattern common to LL parsers \cite{operatorPrecedence}.
Typically, a grammar defines a series of nonterminals - each corresponding to a distinct operator-precedence level (from primaries up to full expressions) - and structures its productions so as to eliminate left recursion while preserving left-associative evaluation at every level.
This arrangement guarantees that the parser can always choose the correct production with a single lookahead token, ensuring LL(1) compliance.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos, breaklines]{yaml}
ProbChoice:
    '{'numerator=Expression ':' denominator=Expression '}';
Expression:
    LogicalOr ({infer ProbabilisticAssignment.head=current} (probabilities+=(ProbChoice) fallbacks+=LogicalOr)+)?;
LogicalOr infers Expression:
.
.
.
Primary infers Expression:
    {infer BoolLiteral} literal=BOOL | 
    {infer IntLiteral} literal=IntegerLiteral |
    Lval |
    '(' Expression ')';
\end{minted}
\captionof{listing}{Langium grammar expression system of MiniProb}

@@

The excerpt displayed here, defines the types alongside its literal lexical vocabulary. The nonterminal \code{Type} accepts either the literal keyword \code{bool} or an \code{IntType}. An \code{IntType} specifies if it's signed and of what width and may be converted to an \code{IntArray} if brackets are present (line 4).
The \code{IntegerLiteral} rule recognizes numeric constants with an optional printable sign. Together, these productions ensure that boolean types, integers, and arrays are adequately represented according to the \code{MiniProb} definition (\ref{sec:miniprob}).

@@

terminal INT_PREFIX: /[su][1-9][0-9]{0,8}/;
terminal BOOL returns boolean: /(true|false)/;
.
terminal ID: /(?!(true|false)|[su][1-9][0-9]*)[a-zA-Z_][a-zA-Z0-9_\.\:\~]*/;
terminal STRING returns string: /"(\\.|[^"\\])*"|'(\\.|[^'\\])*'/;
terminal INT returns number: /[0-9]+/;
hidden terminal WS: /\s+/;
hidden terminal ML_COMMENT: /\/\*[\s\S]*?\*\//;
hidden terminal SL_COMMENT: /\/\/[^\n\r]*/;
\end{minted}
\captionof{listing}{Terminals and patterns of MiniProb}

@@

Langium offers the option of formalizing semantic checks and type systems through a static validation service which can be registered in its dependency-injection framework.
Once registered, this service traverses the abstract syntax tree via a visitor pattern, invoking node-type specific validation routines at each step.
The following sections describe the implementation of static semantics and the design of the type system.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos]{ts}
  export function registerValidationChecks(services: MiniProbServices) {
  const registry = services.validation.ValidationRegistry;
  const validator = services.validation.MiniProbValidator;
  const checks: ValidationChecks<MiniProbAstType> = {
    Lval: validator.checkArrayAccess,
    FuncCall: validator.checkFunctionCalls,
    Func: validator.checkFunctionDefinitions,
    .
    .
    IntegerLiteral: validator.checkIntegerLiteral,
    Program: validator.checkMainOccurrences
  };
  registry.register(checks, validator);
}
\end{minted}
\captionof{listing}{Validation checks registrations}

These validations are AST-driven logical procedures that verify specified conditions on the syntax tree. This means that validation is performed by statically inspecting
each node's structure, attributes, and inter-node relationships, without relying on runtime information or any other states.
Each procedure either invokes the \code{ValidationAcceptor} to report deviations when constraints are violated or lets the node pass unremarked when it
satisfies the criteria. As an example, consider the following \code{checkMainOccurrences} method, which enforces the invariant that a \code{main()} function must be defined exactly once in each program.
\begin{minted}[bgcolor=lightgray, frame=lines, linenos]{ts}
  checkMainOccurrences(node: Program, accept: ValidationAcceptor) {
     const functionNames = node.functions.filter(f => f.name === "main")
            .map((f) => f.name);
     if (functionNames.length <= 0) {
      accept("error", 'Each program needs a \'main()\' function.', {
        node,
        property: "functions"
      });
     }
     .
     .
}
\end{minted}
\captionof{listing}{Sample of validation process}
\vspace{1em}

@@

The cache is instantiated anew for each inference run on a node, eliminating any risk of residual data or cross-contamination and ensuring fully deterministic results.
However, because it isn't persisted between runs, identical subtrees get re-evaluated for every node, which may introduce redundant work.
Unlike incremental parsing systems that retain state to avoid reprocessing unchanged regions, this approach performs a complete, conventional traversal each time.
??try the cache service and see if better???
\\
\\
As an example of how the type system is employed in the validator, this code snippet creates a new cache (a map of \code{ASTNode} -> \code{TypeDescription}) and calls the
\code{inferType} method, which returns the concrete type of the operand. If any inference steps conflict during this process, an \code{ErrorType} is returned and the
propagated message is displayed. Otherwise, the validator verifies that the resolved type is allowed for this operand and handles the error accordingly.

@@

    if (isErrorType(operandType)) {
      accept('error', operandType.message, { node: operandType.source ?? node });
      return;
    }
    if (!isLegalOperation(node.operator, operandType)) {
      accept(
        'error',
        `The operation '${node.operator}' is not possible on type '${typeToString(operandType)}'`,
        { node, property: 'operand' }
      );
    }
  }
\end{minted}
\captionof{listing}{Unary expression validation checks}

@@

The custom scope provider focuses specifically on two kinds of nodes - function calls (\code{FuncCall}) and variable references (\code{LVal}) - since
these are the only AST nodes in the MiniProb grammar that establish referential links to other symbols. By subclassing the base \code{ScopeProvider} and overriding
its \code{getScope} method, the provider can dispatch to specialized logic for these two cases while delegating all other node types back to the default implementation.
This fallback invocation of \code{super.getScope()} outside of the \code{FuncCall} and \code{LVal} branches ensures that any other or future nodes will receive a fully defined and well-formed scope, preserving completeness of the overall resolution process.

Specialized handling is required because the language supports an import mechanism that selectively brings in only top-level definitions from other files, while excluding
all other symbols, which are exported by default into a workspace-wide scope governed by the \code{IndexManager}. In the \code{LVal} branch, the provider correctly resolves
multiple identifiers declared in a single declarations statement (for example, \code{u8 i1, i2, i3}) distinguishing between block-scope and global variables, as well as including function parameters as valid resolutions.
In the \code{FuncCall} branch, default behavior is overridden to prevent 'ghost' functions - placeholder symbols automatically invented for unresolved identifiers - and to include in-flight local declarations that appear before their closing delimiters in live editing scenarios (a solution which inspects raw \code{CST} text and may require future revision).

Because both \code{FuncCall} and \code{LVal} reference the same identifier token pattern, \code{getScope} is invoked for both branches on the same node.
If distinguishing syntactic context - such as surrounding parentheses for calls or the presence of an equals sign for assignments - is present, the provider resolves only once aware of what path to choose.
\\

@@

    const importKey = 'imported-';
    if (targetNodeType === Func) {
      const importedFuncDescriptions = this.descriptionCache.get(
        currentUri,
        importKey + targetNodeType,
        () => new MapScope(this.indexManager.allElements(Func, new Set<string>(importUris)))
      );
      return importedFuncDescriptions.getAllElements();
    } 
    .
    .
  }
\end{minted}
\captionof{listing}{Inclusion of imported file's top-level scope}
\vspace{0.3em}

In this snippet, the overridden \code{getScope} method first confirms that the current resolution request pertains to any \code{reference} context by checking
\code{context.property} and container presence (lines 1-3), then locates the enclosing program node. Upon detecting a function-call node (line 11), it filters the program's
function list to remove placeholder or 'ghost' entries (line 7). A local cache - persistent throughout the whole linking phase - is
then used to retrieve or populate descriptors for these real functions (lines 11-22), encapsulating them within Langium's offered scope structure. If file-level imports are present
(lines 8, 26-30), a helper routine leverages the global \code{IndexManager} to fetch additional symbols from imported modules, yielding a secondary stream of descriptors.
The in-file and imported descriptors, together with any local declarations, are merged into a composite scope (lines 32-34) that is returned for code completion and reference
linking; all other situations fall back to \code{super.getScope()} (lines 38), preserving the default resolution behavior.

@@

        return new MapScope(
          stream(descriptions, importedDescriptions, localDeclarationsDescriptions)
        );
      } 
      .
      .
    return super.getScope(context);
}}
\end{minted}

\section{VS Code Extension Points}
textmate highlighting lang-config.json

??maybe ref from the firt mention of the extension (structure)

@@

\section{Parsing Speed}

>>>>

As discussed in in the introduction \ref{sec:scope}, the main aim is the development of a VS Code Language extension that operates similarly to
well-established tooling for other programming languages. The following requirements define the expected capabilities and characteristics the extension must demonstrate.

@@

  \subsection*{Non-Functional Requirements}
  \begin{enumerate}
    \item The extension must respond to user edits with syntax and semantic feedback within 100-250ms for files of average size.
    \item The extension must provide a user-friendly experience that is consistent with other popular VS Code extensions.
    \item The extension's code must be structured, commented and documented in a way that allows future developers to understand and extend its functionality easily.
    \item The extension must be compatible with the latest version of Visual Studio Code at the time of its release, with consideration for future maintenance and version support.
    \item The extension must be able to handle files of immense size (around 10000 lines) of code within the bounds of 1 second.
    \item The extension must be delivered with clear and comprehensive documentation, including installation, usage, and extension of its features.
  \end{enumerate}
\end{multicols}

\section{Familiar Technologies}

This extension is implemented using \textbf{Langium}, which itself builds upon the standardized \textbf{Language Server Protocol (LSP)} \cite{LSP}, yet
there are several other technologies and approaches available for creating Visual Studio Code language extensions. At its core, the LSP allows tooling for programming
languages to be developed in a modular, editor-independent manner, making it possible to implement a language server, communicating the
language specific support via JSON-RPC.

@@

Meanwhile, incremental parsing frameworks such as \textbf{Tree-sitter}~\cite{Treesitter} have gained popularity for their ability to provide high-performance syntax parsing and error recovery,
making them ideal for tooling that operates on very large files.

For basic syntax highlighting and lexical differentiation, \textbf{TextMate}-style grammars \cite{TextMate} can be used - for editors supporting it, offering a lightweight approach.
An alternative are Monaco-based editors, that employ \textbf{Monarch} typescript files to define the token coloring, in contrast to TextMate which uses JSON files.

@@

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as
    the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
    from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
    All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
    which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textsf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
    dependency-injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
    tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
    state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
    Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possi ble. Moreover, Langium enables developers to swap in
    custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec:customservices}.
    By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
    However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

Both \code{module} files share analogous responsibilities by encapsulating dependency-injection setup and service registration for the language environment.
The generated module is produced automatically from the grammar definition and offers two DI fragments, one that registers shared-core services such as AST
reflection and another that configures the parser, the Chevrotain grammar factory and language metadata. In contrast the main module has to be manually extended and imports
Langium's default core and shared services, registers any custom services such as validators, scope providers and caches, invokes \code{inject} to assemble the full
service container and exposes a \code{createServices} factory. The shared services differentiate themselves in that they can be used cross-language and are not bound to
specific implementations.

@@

\subsection*{Language Server Protocol: Server-Client interaction}

This illustration is a comprehension aid to show how the Language Server Protocol handles standardized communication between an editor (client) and a language server:
from the initialization handshake through document events, feature requests (completion, hover, definition, references, formatting, semantic tokens), and shutdown \cite{lspFeatures}.
By defining a common set of well-specified messages and payloads, LSP serves as a generally applicable protocol that any editor and any language server can adopt to instantly
share rich tooling capabilities without per-language or per-IDE custom integrations.

@@

A Langium grammar definition primarily consists of rules: \code{<Rulename>: <production_result>} and terminals: \code{terminal <name>: <result>} defining the language
in accordance of conventional grammatical specifications. In addition, it must also include keywords specifying the grammar name also used in the AST: \code{grammar <name>},
and to determine the start from which a document should be parsed: \code{entry <RootRulename>: <production_result>}.
In this case, the production result describes the whole parsable document and therefore acts as the root node of the AST.
\\
The code excerpts in the following sections are all taken from the original MiniProb Langium grammar (\ref{list:langium-grammar}) and, for easier digestion,
have been truncated; each snippet focuses respectively on the core rules, the expression system, and the terminal and type definitions.
\subsection*{Core rules}

The code excerpt shown below defines 'MiniProb' as the grammar name and sets the \code{Program} rule as the entry point, adhering to nominal abstraction by
referring to each MiniProb-parsed document as a program. Lines of interest include line 8 - showing that every MiniProb program consists of some global declarations
followed by at least on function - and line 4, which allows for file import of other MiniProb programs at the beginning of the document.

The \code{FileImport} rule emulates the C preprocessor's "\#include" directive by requiring a literal match of the \texttt{\#include} keyword followed by a
file name token (terminal \code{STRING}). Generally during parsing, each line is evaluated either against these fixed string keywords or against any matching terminals (order of match).
As the file names have to be retrievable once it is time to compute the scope \ref{sec:customservices}, they are made accessible through the AST
, with each FileImport-node containing a \code{file} property that stores the corresponding value.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos, breaklines]{yaml}
grammar MiniProb
type DeclOrParam = Decl | Param;
entry Program:
    (fileImports+=FileImport)*
    probabilisticQuery=PROB_QUERY? // not tested: out of scope
    formula=FORMULA? // not tested: out of scope
    'program:'?
    (declarations+=Decl ';')* functions+=Func+;
FileImport:
    '#include' file=STRING;
Decl:
    type=Type names+=ID (',' names+=ID)*;
.
.
Lval:
    ref=[DeclOrParam:ID] ('['index=Expression']')?;
.
.
Param:
    type=Type byRef?='&'? name=ID;
ArgList:
    arguments+=Arg (',' arguments+=Arg)*;
Arg:
    expression=Expression; //only expression and no by-ref symbol (could be replaced)
.
.
\end{minted}
\captionof{listing}{Langium grammar core of MiniProb}
\vspace{0.5em}

@@

Most of the expression system is omitted here, as it closely follows the \textit{operator-precedence grammar} pattern common to LL parsers.
Typically, a grammar defines a series of nonterminals - each corresponding to a distinct operator-precedence level (from primaries up to full expressions) - and structures its productions so as to eliminate left recursion while preserving left-associative evaluation at every level.
This arrangement guarantees that the parser can always choose the correct production with a single lookahead token, ensuring LL(1) compliance.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos, breaklines]{yaml}
ProbChoice:
    '{'numerator=Expression ':' denominator=Expression '}';
Expression:
    LogicalOr ({infer ProbabilisticAssignment.head=current} (probabilities+=(ProbChoice) fallbacks+=LogicalOr)+)?;
LogicalOr infers Expression:
.
.
.
Primary infers Expression:
    {infer BoolLiteral} literal=BOOL | 
    {infer IntLiteral} literal=IntegerLiteral |
    Lval |
    '(' Expression ')';
\end{minted}
\captionof{listing}{Langium grammar expression system of MiniProb}
\vspace{0.5em}

@@

The excerpt displayed below, defines the types alongside its literal lexical vocabulary. The nonterminal \code{Type} accepts either the literal keyword \code{bool} or an \code{IntType}. An \code{IntType} specifies if it's signed and of what width and may be converted to an \code{IntArray} if brackets are present (line 4).
The \code{IntegerLiteral} rule recognizes numeric constants with an optional printable sign. Together, these productions ensure that boolean types, integers, and arrays are adequately represented according to the \code{MiniProb} definition (\ref{sec:miniprob}).

@@

terminal INT_PREFIX: /[su][1-9][0-9]{0,8}/;
terminal BOOL returns boolean: /(true|false)/;
.
terminal ID: /(?!(true|false)|[su][1-9][0-9]*)[a-zA-Z_][a-zA-Z0-9_\.\:\~]*/;
terminal STRING returns string: /"(\\.|[^"\\])*"|'(\\.|[^'\\])*'/;
terminal INT returns number: /[0-9]+/;
hidden terminal WS: /\s+/;
hidden terminal ML_COMMENT: /\/\*[\s\S]*?\*\//;
hidden terminal SL_COMMENT: /\/\/[^\n\r]*/;
\end{minted}
\captionof{listing}{Terminals and patterns of MiniProb}
\vspace{0.5em}

@@

Langium offers the option of formalizing semantic checks and type systems through a static validation service which can be registered in its dependency-injection framework.
Once registered, this service traverses the abstract syntax tree, invoking node-type specific validation routines at each step.
The following sections describe the implementation of static semantics and the design of the type system.

@@

\begin{minted}[bgcolor=lightgray, frame=lines, linenos]{ts}
  export function registerValidationChecks(services: MiniProbServices) {
  const registry = services.validation.ValidationRegistry;
  const validator = services.validation.MiniProbValidator;
  const checks: ValidationChecks<MiniProbAstType> = {
    Lval: validator.checkArrayAccess,
    FuncCall: validator.checkFunctionCalls,
    Func: validator.checkFunctionDefinitions,
    .
    .
    IntegerLiteral: validator.checkIntegerLiteral,
    Program: validator.checkMainOccurrences
  };
  registry.register(checks, validator);
}
\end{minted}
\captionof{listing}{Validation checks registrations}
\vspace{0.5em}

These validations are AST-driven logical procedures that verify specified conditions on the syntax tree. This means that validation is performed by statically inspecting
each node's structure, attributes, and inter-node relationships, without relying on runtime information or any other states.
Each procedure either invokes the \code{ValidationAcceptor} to report deviations when constraints are violated or lets the node pass unremarked when it
satisfies the criteria. As an example, consider the following \code{checkMainOccurrences} method, which enforces the invariant that a \code{main()} function must be defined exactly once in each program.
\begin{minted}[bgcolor=lightgray, frame=lines, linenos]{ts}
  checkMainOccurrences(node: Program, accept: ValidationAcceptor) {
     const functionNames = node.functions.filter(f => f.name === "main")
            .map((f) => f.name);
     if (functionNames.length <= 0) {
      accept("error", 'Each program needs a \'main()\' function.', {
        node,
        property: "functions"
      });
     }
     .
     .
}
\end{minted}
\captionof{listing}{Sample of validation process}
\vspace{0.5em}

@@

The cache is instantiated anew for each inference run on a node, eliminating any risk of residual data or cross-contamination and ensuring fully deterministic results.
However, because it isn't persisted between runs, identical subtrees get re-evaluated for every node, which may introduce redundant work.
Unlike incremental parsing systems that retain state to avoid reprocessing unchanged regions, this approach performs a complete, conventional traversal each time.
\\
\\
As an example of how the type system is employed in the validator, this code snippet creates a new cache (a map of \code{ASTNode} -> \code{TypeDescription}) and calls the
\code{inferType} method, which returns the concrete type of the operand. If any inference steps conflict during this process, an \code{ErrorType} is returned and the
propagated message is displayed. Otherwise, the validator verifies that the resolved type is allowed for this operand and handles the error accordingly.

@@

    if (isErrorType(operandType)) {
      accept('error', operandType.message, { node: operandType.source ?? node });
      return;
    }
    if (!isLegalOperation(node.operator, operandType)) {
      accept(
        'error',
        `The operation '${node.operator}' is not possible on type '${typeToString(operandType)}'`,
        { node, property: 'operand' }
      );
    }
  }
\end{minted}
\captionof{listing}{Unary expression validation checks}
\vspace{0.5em}

@@

The custom scope provider focuses specifically on two kinds of nodes - function calls (\code{FuncCall}) and variable references (\code{LVal}) - since
these are the only AST nodes in the MiniProb grammar that establish referential links to other symbols. By subclassing the base \code{ScopeProvider} and overriding
its \code{getScope} method, the provider can dispatch to specialized logic for these two cases while delegating all other node types back to the default implementation.
This fallback invocation of \code{super.getScope()} outside of the \code{FuncCall} and \code{Lval} branches ensures that any other or future nodes will receive a fully defined and well-formed scope, preserving completeness of the overall resolution process.

Specialized handling is required because the language supports an import mechanism that selectively brings in only top-level definitions from other files, while excluding
all other symbols, which are exported by default into a workspace-wide scope governed by the \code{IndexManager}. In the \code{Lval} branch, the provider correctly resolves
multiple identifiers declared in a single declarations statement (for example, \code{u8 i1, i2, i3}) distinguishing between block-scope and global variables, as well as including function parameters as valid resolutions.
In the \code{FuncCall} branch, default behavior is overridden to prevent 'ghost' functions - placeholder symbols automatically invented for unresolved identifiers - and to include in-flight local declarations
that appear before their closing delimiter in live editing scenarios (a solution which inspects raw \code{CST} text and may require future revision).

Because both \code{FuncCall} and \code{Lval} reference the same identifier token pattern, \code{getScope} is invoked for both branches on the same node.
If distinguishing syntactic context - such as surrounding parentheses for calls or the presence of an equals sign for assignments - is present, the provider resolves only once, aware of what path to choose.
\\

@@

    const importKey = 'imported-';
    if (targetNodeType === Func) {
      const importedFuncDescriptions = this.descriptionCache.get(
        currentUri,
        importKey + targetNodeType,
        () => new MapScope(this.indexManager.allElements(Func, new Set<string>(importUris)))
      );
      return importedFuncDescriptions.getAllElements();
    } 
    .
    .
  }
\end{minted}
\captionof{listing}{Inclusion of imported file's top-level scope}
\vspace{0.5em}

In the snippet below, the overridden \code{getScope} method first confirms that the current resolution request pertains to any \code{reference} context by checking
\code{context.property} and container presence (lines 1-3), then locates the enclosing program node. Upon detecting a function-call node (line 11), it filters the program's
function list to remove placeholder or 'ghost' entries (line 7). A local cache - persistent throughout the whole linking phase - is
then used to retrieve or populate descriptors for the real functions (lines 11-22), encapsulating them within Langium's offered scope structure. If file-level imports are present
(lines 8, 26-30), a helper routine leverages the global \code{IndexManager} to fetch additional symbols from imported modules, yielding a secondary stream of descriptors.
The in-file and imported descriptors, together with any local declarations, are merged into a composite scope (lines 32-34) that is returned for code completion and reference
linking; all other situations fall back to \code{super.getScope()} (lines 38), preserving the default resolution behavior.

@@

        return new MapScope(
          stream(descriptions, importedDescriptions, localDeclarationsDescriptions)
        );
      } 
      .
      .
    return super.getScope(context);
}}
\end{minted}
\captionof{listing}{Final scope formation and provision}
\vspace{0.5em}

\section{VS Code Extension Points}

@@

\chapter{Conclusion}

reqjuiremnts fullfilled, code completeion lacks in basic stucutrues only for linked references.
conlsuio: mention no dode comülpeteion for structures only refs, not optimal and how it would be implemetned
